---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Zain Naboulsi"
date: "2023-05-23"
bibliography: references.bib
output:
  html_document:
    toc: yes
    df_print: paged
    
  pdf_document:
    toc: yes
    includes:
      in_header: header.tex

editor_options: 
  markdown: 
    wrap: 80
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}
# set a seed in case we use any random numbers
set.seed(1337)


# Set the names of the packages and libraries you want to install
required_libraries <- c("tidyverse", "purrr", "knitr", "skimr", "readr")

# Install missing packages and load all required libraries
for (lib in required_libraries) {
  if (!requireNamespace(lib, quietly = TRUE)) {
    install.packages(lib)
  }
  library(lib, character.only = TRUE)
}

# load all the cycling data for analysis
# our data runs from 04/2022 to 04/2023
if(file.exists("combined_data_df.rds")) {
    # read in our combined data
    combined_data_df <- readRDS("combined_data_df.rds")
    
} else {
    
    # get file list
    file_list <- list.files(path = "SourceData/", pattern = "*.csv", full.names = TRUE)
    
    # read and combine files
    combined_data_df <- map_df(file_list, read_csv)
    
    # convert ridable_type and member_casual into factors
    combined_data_df$rideable_type <- as.factor(combined_data_df$rideable_type)
    combined_data_df$member_casual <- as.factor(combined_data_df$member_casual)
    
    # save the combined_data object so we don't have to reload this data again
    # unless it changes
    write_rds(combined_data_df, "combined_data_df.rds")
}

```

# Abstract
Cyclistic, a popular bike-share company based in Chicago, strives to maximize 
its annual memberships for sustainable growth. The company has an existing user 
base composed of both casual riders (who opt for single-ride or full-day passes) 
and annual members. The focus of this case study is to understand the 
utilization patterns of these two user groups, as the company believes that 
converting casual riders into annual members can significantly enhance their 
profitability. To devise effective marketing strategies for this conversion, the 
case study will delve into questions that probe how these two user groups use 
Cyclistic bikes differently, why casual riders might opt for annual memberships, 
and how digital media can facilitate this conversion.


# Primary Business Question  
The main business question we seek to address is: 

**How do annual members and casual riders use Cyclistic bikes differently?**  

To answer these questions, we will delve into the following topics:

1. *Duration and frequency of rides:* We will analyze data to understand the 
typical length of rides for both user groups. Are there noticeable 
differences between casual riders and members? Are members using the bikes more 
frequently or for longer duration?

2. *Ride timings:* We'll investigate if there's a difference in the time of day, 
day of the week, or month of the year when the two user groups typically use the 
bikes.

3. *Bike preferences:* Given that Cyclistic offers various bike options, we will 
examine if there's a pattern in the type of bike preferred by the two user 
groups.

Answering these questions will provide insights into the riding patterns of the 
two groups (annual member and casual renters), and will be instrumental in 
forming the foundation for a marketing strategy aimed at converting casual 
riders into annual members. The strategy will be developed in alignment with 
these insights, addressing the unique needs and behaviors of casual riders to 
motivate their transition to becoming annual members.

Cyclistic's collected data forms the backbone of this analysis. For 
transparency, we have included the source code in the paper's Appendix. We 
strive to communicate our findings clearly, pairing concise language with 
comprehensible visualizations and straightforward recommendations.

# About the Data  

The data for this analysis is publicly available and located on Divvy's 
(aka Cyclistic) Amazon S3 storage. The data is titled as 
"Divvy Data." [@DivvyData2023]

The data is organized in an anonymized format, detailing each trip's start and 
end day/time, start and end stations, and rider type (Member or Casual). Trips 
taken by staff and any trips under 60 seconds in length have been excluded from 
this dataset.

Given the anonymization and rigorous data handling practices outlined by Divvy, 
we can consider the data to be Reliable, Original, Comprehensive, Current, and 
Cited (ROCCC). However, some level of bias could be inherent as the data only 
includes riders who choose to use Divvy's services, and not all potential or 
actual cyclists in Chicago. Also, the credibility of the dataset is tied to 
Divvy's data collection and handling processes.

The data is provided according to the Divvy Data License Agreement. 
Accessibility is facilitated through public download links, ensuring the data 
can be freely used for analysis while respecting privacy through anonymization. 
Security of the data in our analysis will be ensured by following best 
practices, such as secure data storage and handling.

Data integrity has been verified through initial exploratory data analysis to 
check for missing values, inconsistencies, or obvious outliers. Any issues found 
during this process will be documented and addressed appropriately.

This dataset is instrumental in answering our primary business question — 
understanding how annual members and casual riders use Cyclistic bikes 
differently. Each trip's information, coupled with the type of rider, will offer 
insights into the usage patterns of the two user groups.

Potential problems with the data could include missing values, inconsistent data 
entries, or anomalies. These will be explored and addressed during the data 
cleaning and preprocessing stages of our analysis. If severe issues are 
encountered, they will be documented and taken into consideration when 
interpreting the analysis results.

## Structure of the Data
We used the str() function in R to give us a detailed structure of the data.  

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# examine the structure of the data
str(combined_data_df)

```

Here are some interesting observations from the output:  

* __Data size:__ The dataframe contains over 6.23 million records, spread across 13 
variables.  

* __Data types:__ The variables in the dataframe are of various types, including 
character (e.g., ride_id, start_station_name, start_station_id, etc.), 
factor (e.g., rideable_type, member_casual), numeric (e.g., start_lat, 
start_lng, end_lat, end_lng), and POSIXct (e.g., started_at, ended_at). This 
variety indicates that the dataset is rich and contains different kinds of 
information.  

* __Rideable types:__ There are three levels of rideable_type: classic_bike, 
docked_bike, and electric_bike.  

* __Member categories:__ There are two levels of member_casual: casual and member.  

* __Geographic coordinates:__ The data contains latitude and longitude information 
for both the start and end locations of rides. This information could be used to 
plot geographic routes or to analyze usage patterns across different locations.  

* __Time information:__ The dataset includes when each ride started and ended. This 
information can be used to analyze the temporal patterns of bike usage.

* __Station information:__ The data includes detailed information about the start 
and end stations, including their names and IDs.

## Duplicate Observations
Now we turn our attention to broad issues in the data if they exist. The first 
task we will undertake is to check for duplicate rows in our data. Fortunately, 
this dataset doesn't have any duplicate data. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Identify duplicates
duplicates <- duplicated(combined_data_df)

# Count duplicates
num_duplicates <- sum(duplicates)

# Show the number of duplicates
cat("Total number of duplicate rows found:",num_duplicates)

```  



## Missing Values
Our next task is to look for missing data (or NA data, as it is commonly 
called). Here is what we found:

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Check for missing values
colSums(is.na(combined_data_df))

```

From this output, we can see that there are missing values (NA) in some of the 
columns.

- The `start_station_name` column has 902,896 missing values.
- The `start_station_id` column has 903,028 missing values.
- The `end_station_name` column has 964,949 missing values.
- The `end_station_id` column has 965,090 missing values.
- The `end_lat` and `end_lng` columns each have 6,290 missing values.

The other columns (`ride_id`, `rideable_type`, `started_at`, `ended_at`, 
`start_lat`, `start_lng`, and `member_casual`) do not have any missing values.

Missing data can be problematic because it may skew the results of our analysis 
or cause certain functions to fail. You will want to investigate why there are 
missing values in the data and consider ways of handling it, such as filling 
them in with an average or other statistic (imputation), or removing rows. 

The differences in the number of missing values between the `start` and `end` 
station names and IDs could indicate inconsistencies in data collection or 
recording. This may require further investigation to ensure that any analyses 
we conduct on this data are valid.


## Summary Statistics
Now we an run some summary statistics to get an overall view of the data 
makeup and how we might approach our analysis. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Print detailed summary statistics of the dataframe
skim(combined_data_df)

```

Here are some insights we can glean:

* __Data dimensions:__ The dataset contains 6,230,310 rows (observations) and 
13 columns (variables). It's a large dataset.

* __Variable types:__ There are five character variables, two factors, 
four numeric variables, and two POSIXct (date-time) variables. This mix of types 
suggests a varied dataset, likely to contain both categorical and continuous 
data, as well as time series data.

* __Missing Values:__ The start and end station names and IDs have around 15% 
missing values, given that the complete rate is around 0.85. This could be a 
potential area of concern depending on how we want to use this data.

* __Unique Values:__ The ride_id variable is unique for every row, which 
correctly suggests that it is a unique identifier for each ride.

* __Rideable Type:__ There are three types of bikes used in these rides: 
electric bikes, classic bikes, and docked bikes. The most common type of bike 
used is the electric bike, followed by the classic bike.

* __Membership:__ There are more rides by members (3,745,586) than by casual 
users (2,484,724).

* __Locations:__ The latitude and longitude for start and end points of the 
rides are quite concentrated around certain values (41.9 for latitude and -87.6 
for longitude), suggesting that most rides occur in a relatively confined 
geographical area.

* __Dates:__ The started_at and ended_at fields indicate that the data spans 
from 1st April 2022 to 30th April 2023, so just over a year's worth of ride 
data.


# Processing the Data
Turning our attention to processing the data, we can review what decisions we 
have made thus far and augment the data as needed to facilitate the analysis 
process coming up. 

## Key Decisions
Here are the decisions made so far:
* __Tools chosen__ include R programming language along with the `tidyverse`, 
`purrr`, `knitr`, `skimr`, and `readr` libraries. The reason for choosing these tools is 
their strength in data manipulation, visualization, and analysis. `tidyverse` 
includes several useful packages for data cleaning and analysis, while `purrr` 
enhances R's functional programming capabilities. `knitr` provides dynamic 
report generation, and `skimr` and `readr` are helpful for data summarization 
and importing.

* __Data's integrity__ is ensured through multiple steps: Firstly, by loading 
the dataset directly from a local ".rds" file if it exists to avoid data 
corruption that could occur during multiple read operations. If the ".rds" file 
does not exist, CSV files are read directly from the source directory, ensuring 
the original data's integrity.

* __Data cleaning__ steps taken involve checking for missing values and 
duplicated records. The number of missing values in each column is found using 
`colSums(is.na(combined_data_df))`, and duplicated records are found using 
`duplicated(combined_data_df)`.

* __Data verification__ for readiness for analysis is done by summarizing the 
dataframe using `skim(combined_data_df)` which provides detailed summary 
statistics. This summary provides a good understanding of the data's 
distribution and potential issues (like extreme outliers).

* __Cleaning processes__ are documented in this markdown document. Including, 
checking for missing values, identifying duplicates, and getting summary 
statistics. This document serves as a record of the cleaning process and can be 
shared or reviewed as needed.

## Additional Data Changes  
Beyond the standard processing steps to get the data ready for analysis, we 
decided to make a couple of modifications to the dataframe to support additional 
queries we anticipate. The changes are as follows:  

* __Calculate Trip Time:__ The `ride_length` variable was added. This column is a 
derived by subtracting `ended_at` from `started_from`. The results are in total 
minutes elapsed for the ride.  

* __Start Date:__ One new column (`started_date`) was added to strip out the 
time from the original `started_at` column. This allows for plots to examine 
rental trends over time by day. 

* __Day of Week:__ A two new columns entitled `day_of_week_start` and 
`day_of_week_end` were added to make analysis of the specific day (e.g. Monday, 
Tuesday, etc...) a trip happened easier. This is converted to a factor for more 
advanced activities.  


```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# calculate the total ride time in minutes for each group
combined_data_df$ride_length <- 
    round((combined_data_df$ended_at - combined_data_df$started_at) / 60, 2)

# Change the units attribute to "minutes"
attr(combined_data_df$ride_length, "units") <- "minutes"


# Add a new column for the date without the time
combined_data_df$started_date <- as.Date(combined_data_df$started_at)


# Create the 'day_of_week_start' and 'day_of_week_end' columns
combined_data_df$day_of_week_start <- wday(combined_data_df$started_at, label = TRUE)
combined_data_df$day_of_week_end <- wday(combined_data_df$ended_at, label = TRUE)

# Convert the new columns to factors
combined_data_df$day_of_week_start <- as.factor(combined_data_df$day_of_week_start)
combined_data_df$day_of_week_end <- as.factor(combined_data_df$day_of_week_end)


```

## Removing Outliers
After adding the ride_length variable we discovered some outliers in the data. 
Specifically, we found start and end trip times that would result in a negative 
(or very low positive) time and we found values that were well beyond the two day "normal" time for a 
ride. Ultimately, we decided to limit the results to those rentals that were 
between five minutes and two days long. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Get rid of outliers by only accepting ride times that are less than two days 
# and greater than zero
combined_data_df <- combined_data_df %>% filter(ride_length <= 2880 & 
                                                    ride_length > 5) 


```


# Analysis of the Data

## Rental Basic Metrics 
Overall, it appears that casual customers rent bikes for longer periods of time 
compared to members. This information can be valuable for our strategic 
planning. For instance, it might be worth exploring if there are ways to 
encourage casual customers to become members, given their high usage. It's also 
important to explore the reasons behind the high maximum ride lengths, 
particularly for casual customers, as this might point to potential misuse or 
misunderstanding of rental policies. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# split our data into two groups: member and casual
member_df <- combined_data_df %>% filter(member_casual == "member") 
casual_df <- combined_data_df %>% filter(member_casual == "casual") 

# ride_length metrics for both groups
ride_length_mean_member <- mean(as.numeric(member_df$ride_length))
ride_length_max_member <- max(as.numeric(member_df$ride_length))
ride_length_min_member <- min(as.numeric(member_df$ride_length))

ride_length_mean_casual <- mean(as.numeric(casual_df$ride_length))
ride_length_max_casual <- max(as.numeric(casual_df$ride_length))
ride_length_min_casual <- min(as.numeric(casual_df$ride_length))

# Create a data frame
ride_length_metrics <- data.frame(
    Metric = c("Mean", "Max", "Min"),
    Member = round(c(ride_length_mean_member, ride_length_max_member, ride_length_min_member), 2),
    Casual = round(c(ride_length_mean_casual, ride_length_max_casual, ride_length_min_casual), 2)
)

# Print the data frame
print(ride_length_metrics)


```


## Daily Bike Rentals Over Time  
Now let's take a step back and look at the entire year of data we have 
to get some perspective on rental trends between the two groups. The graph below 
shows the daily bike rentals from April 2022 to April 2023. 

As we can see, there is a definitive seasonal trend to bike rentals overall and 
particularly between the two groups. Most notably, the member rentals decline in 
the winter months but almost never "bottom out". Whereas, the casual rentals go 
very flat in the winter months but make up for it in volume during the warmer 
times. If we are looking to convert casual riders to members, an incentive that 
can be used during the warmer months should be considered. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Group by date and count the number of rides each day per group
daily_counts_member <- member_df %>% group_by(started_date) %>% summarise(count = n())
daily_counts_casual <- casual_df %>% group_by(started_date) %>% summarise(count = n())


# super line graph for members
ggplot(daily_counts_member, aes(x = started_date, y = count)) +
    geom_line(color = "darkblue") +
    geom_smooth(method = "loess", se = FALSE, color = "red", linetype="dashed") +
    labs(x = "Date", 
         y = "Number of Rentals", 
         title = "Daily Bike Rentals for Members", 
         subtitle = "With trend line",
         caption = "Source: https://divvybikes.com/system-data") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20),
          plot.subtitle = element_text(face = "italic", size = 12),
          axis.title = element_text(face = "bold", size = 14),
          axis.text = element_text(size = 12))

# super line graph for casual
ggplot(daily_counts_casual, aes(x = started_date, y = count)) +
    geom_line(color = "darkblue") +
    geom_smooth(method = "loess", se = FALSE, color = "red", linetype="dashed") +
    labs(x = "Date", 
         y = "Number of Rentals", 
         title = "Daily Bike Rentals for Casual", 
         subtitle = "With trend line",
         caption = "Source: https://divvybikes.com/system-data") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20),
          plot.subtitle = element_text(face = "italic", size = 12),
          axis.title = element_text(face = "bold", size = 14),
          axis.text = element_text(size = 12))



```


## Bike Preferences Overall  
When we looked at bike preference type, we found that members, marginally, 
prefer classic bikes. The same is not true for casual riders as they tend to 
prefer electric bikes. Perhaps some incentives around electric bike rentals 
should be considered to move casual riders into becoming members. 

NOTE: While the vast majority of bikes are classic or electric, it appears that 
there are still some with the older designation of docked_bike. A docked bike is 
a traditional type of shared bike that needs to be picked up and returned to a 
specific docking station. Docking stations are fixed physical locations 
distributed throughout a city where bikes are stored. When users are finished 
with their ride, they must return the bike to one of these stations and dock it 
correctly for the trip to end.

Cyclistic, which operates in Chicago, uses both docked bikes and "dockless" 
bikes. Dockless bikes, also called free-floating bikes, can be picked up and 
left anywhere within a designated area. They have built-in locks and GPS 
technology that allow the system to keep track of the bikes' locations.

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Calculate the counts
member_counts <- member_df %>% group_by(rideable_type) %>% summarise(count = n())
casual_counts <- casual_df %>% group_by(rideable_type) %>% summarise(count = n())

# Combine the data into one dataframe
bike_pref_df <- rbind(
    mutate(member_counts, user_type = "Member"),
    mutate(casual_counts, user_type = "Casual")
)

# Plot
ggplot(bike_pref_df, aes(x = rideable_type, y = count, fill = user_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label=count), vjust=-0.3, position = position_dodge(0.9)) +
    labs(x = "Bike Type", y = "Count", fill = "User Type", title = "Bike Type Preference by User Type") +
    theme_minimal()


```

## Day and Time Preferences
Now that we know the seasonal differences in rentals, we can turn our attention 
to the day and time when most rentals happen for our two groups. The heat maps 
below show us this information. 

For member rentals, we can clearly see that most rentals happen Monday 
through Thursday. They tend to happen more in the morning and at night. 
Surprisingly, fewer rentals happen in the morning than at night which might 
suggest some riders are taking alternate transport to work and then renting a 
bike at the end of the day to go home. Further study to discover what those 
modes might be would be useful in building a marketing strategy. 

For casual rentals, the data is reversed. Most rentals happen on the weekends 
and the rentals are fairly consistent throughout those days. This suggests that 
casual rentals are being used for leisure and not work transportation. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Extract hour and day of week from started_at for members
member_df <- member_df %>%
    mutate(hour = hour(started_at),
           day = wday(started_at, label = TRUE)) %>%
    group_by(hour, day) %>%
    summarise(freq = n())

# Extract hour and day of week from started_at for casual customers
casual_df <- casual_df %>%
    mutate(hour = hour(started_at),
           day = wday(started_at, label = TRUE)) %>%
    group_by(hour, day) %>%
    summarise(freq = n())



# Create the heatmap for member rentals
ggplot(member_df, aes(x = day, y = hour, fill = freq)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "light blue", high = "dark blue") +
    labs(title = "Member Rentals by Time and Day of Week",
         x = "Day of Week",
         y = "Hour of Day",
         fill = "Frequency") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))



# Create the heatmap for member rentals
ggplot(casual_df, aes(x = day, y = hour, fill = freq)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "light blue", high = "dark blue") +
    labs(title = "Casual Rentals by Time and Day of Week",
         x = "Day of Week",
         y = "Hour of Day",
         fill = "Frequency") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))


```


## Bike Preferences Over Time
Looking at bike preferences over time between the two groups we don't see many 
surprises. However, there is a very notable split in the warmer months where 
members clearly are choosing classic bikes over electric. This could be due to 
a host of reasons, not the least of which is simple supply and demand since this 
is the busiest time of the year for rentals. However, it should be examined to 
ensure there isn't some other casual factor. 

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# Group by date, user type and bike type to get counts
daily_bike_counts <- combined_data_df %>% 
    group_by(started_date, member_casual, rideable_type) %>%
    summarise(count = n()) %>%
    ungroup()

# Now we'll split the data into two dataframes for member and casual
member_daily_bike_counts <- daily_bike_counts %>% 
    filter(member_casual == "member")

casual_daily_bike_counts <- daily_bike_counts %>% 
    filter(member_casual == "casual")

# Define our color palette
color_palette <- c("classic_bike" = "blue", "electric_bike" = "red", "docked_bike" = "green")

# Plot for Members
ggplot(member_daily_bike_counts, aes(x = started_date, y = count, color = rideable_type)) +
    geom_line() +
    scale_color_manual(values = color_palette) +
    labs(x = "Date", y = "Number of Rentals", 
         color = "Bike Type",
         title = "Daily Bike Type Preferences of Members") +
    theme_minimal()

# Plot for Casual users
ggplot(casual_daily_bike_counts, aes(x = started_date, y = count, color = rideable_type)) +
    geom_line() +
    scale_color_manual(values = color_palette) +
    labs(x = "Date", y = "Number of Rentals", 
         color = "Bike Type",
         title = "Daily Bike Type Preferences of Casual Users") +
    theme_minimal()


```

# Conclusion  

Upon examining the bike-sharing data from  Cyclistic, we've unearthed several 
key insights that could steer the company's strategic decision-making, 
especially when it comes to converting casual riders into members:

1. __Seasonality__: We observed a clear pattern of bike rentals across the 
seasons. Both members and casual riders tend to rent more during warmer months 
and less during winter. However, members' usage is steadier across the year, 
while casual riders' rentals significantly drop during winter. Therefore, 
memberships offer a consistent revenue stream across the year.

2. __Rental Duration__: Casual riders appear to rent bikes for longer periods 
than members. This trend could indicate that casual riders embark on longer, 
less hurried trips. Understanding the reasons behind these long rides could 
guide our marketing strategies to convert more casual riders into members.

3. __Bike Preference__: Members seem to prefer classic bikes slightly more than 
casual riders who lean towards electric bikes. Creating incentives related to 
electric bike rentals could persuade casual riders to sign up for a membership.

4. __Day and Time__: Members usually rent bikes on weekdays during commute 
hours, suggesting that they use the service for work or school transportation. 
In contrast, casual riders tend to rent bikes on weekends throughout the day, 
indicating leisure usage. Marketing campaigns can highlight the benefits of 
membership for both commuting and leisure to attract casual riders.

5. __Data Quality__: We noticed some areas where the data quality could be 
improved. Specifically, the absence of data in the start and end station names 
and IDs could limit the depth of our analysis. Enhanced data collection or 
cleaning methods could improve this aspect and provide more valuable data for 
future analysis.

In conclusion, these findings can help Cyclistic design targeted marketing 
strategies to convert casual riders into members. For instance, promoting 
membership benefits for commuting and leisure, offering incentives for electric 
bike rentals, and intensifying marketing efforts during the warm seasons when 
casual usage soars. Improving data collection practices could provide a richer 
data set, which would be a powerful asset for future decision-making.

# Recommendations
Based on our analysis, we offer the following top three recommendations:

1. __Develop Seasonal Marketing Campaigns__: Since bike rentals show a clear 
seasonal trend with casual riders renting more during warmer months, consider 
developing marketing campaigns that align with this trend. Use targeted 
advertising during these periods to highlight the benefits of membership, 
possibly offering seasonal discounts or promotional codes to incentivize casual 
riders to convert into members.

2. __Promote Electric Bikes__: As casual riders show a preference for electric 
bikes, there could be potential in offering membership benefits specifically 
related to these bikes. Special member rates, priority access, or loyalty 
rewards for electric bike usage could appeal to these casual riders and motivate 
them to sign up for membership.

3. __Highlight Membership Benefits for Leisure and Commute__: Given the 
differing usage patterns between members and casual users, consider tailoring 
communication to highlight the benefits of membership for both commuting and 
leisure use. For instance, emphasize the cost-efficiency of membership for 
weekday commuting to work or school, and the convenience and enjoyment for 
weekend leisure rides. This dual-focused approach could resonate with casual 
riders who are using the service for different purposes.

# Further Study
To make predictions about when a casual rider might become a member, we need 
more comprehensive data. This includes tracking individual user behavior over a 
period of time - observing frequency of rides, duration, preferred stations, and 
other relevant features. It is essential to have data where we see casual riders 
transitioning to become members.

With the appropriate data, we could apply techniques such as survival analysis 
or other time-to-event models. These models could potentially predict the timing 
of a casual rider converting to a member.

However, we should recognize the complexity of this task. Numerous factors 
influence the decision of a casual rider to become a member. These factors can 
range from personal financial circumstances, lifestyle choices, and the quality 
of their experiences with the service. Understanding these nuances will require 
a multi-dimensional approach and thoughtful interpretation of data.

Therefore, it's crucial that we broaden our data collection and analysis to 
better understand these dynamics. By doing so, we will be able to create 
accurate models and develop strategic insights to drive membership growth.


\newpage
# Appendix

## References  

<div id="refs"></div>  

## Source Code  
All source code is written in the R Programming Language [@rbase] using the 
R Studio IDE [@rstudio]. The packages used include the popular tidyverse 
[@tidyverse] set of packages for doing most of the heavy lifting in the code, 
knitr [@knitr] for rendering the R Markdown [@rmarkdown] documents, and 
skimr [@skimr] for creating detailed summaries of the data. Finally, for 
generating the visuals, we used the ggplot2 [@ggplot2] package. 


```{r ref.label = knitr::all_labels(appendix == TRUE), echo=TRUE, eval=FALSE}
