---
title: "Title Here"
author: "Zain Naboulsi"
date: "2023-03-22"
bibliography: references.bib
output:
  html_document:
    toc: yes
    
  pdf_document:
    toc: yes
    
editor_options: 
  markdown: 
    wrap: 80
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}
# set a seed in case we use any random numbers
set.seed(1337)


# Set the names of the packages and libraries we want to install
required_libraries <- c("tidyverse", "purrr", "knitr")

# Install missing packages and load all required libraries
for (lib in required_libraries) {
  if (!requireNamespace(lib, quietly = TRUE)) {
    install.packages(lib)
  }
  library(lib, character.only = TRUE)
}

# load all the cycling data for analysis
# our data runs from 04/2022 to 04/2023
if(file.exists("combined_data_df.rds")) {
    # read in our combined data
    combined_data_df <- readRDS("combined_data_df.rds")
    
} else {
    
    # get file list
    file_list <- list.files(path = "SourceData/", pattern = "*.csv", full.names = TRUE)
    
    # read and combine files
    combined_data_df <- map_df(file_list, read_csv)
    
    # convert ridable_type and member_casual into factors
    combined_data_df$rideable_type <- as.factor(combined_data_df$rideable_type)
    combined_data_df$member_casual <- as.factor(combined_data_df$member_casual)
    
    # save the combined_data object so we don't have to reload this data again
    # unless it changes
    write_rds(combined_data_df, "combined_data_df.rds")
}

```

# Abstract
Cyclistic, a popular bike-share company based in Chicago, strives to maximize 
its annual memberships for sustainable growth. The company has an existing user 
base composed of both casual riders (who opt for single-ride or full-day passes) 
and annual members. The focus of this case study is to understand the 
utilization patterns of these two user groups, as the company believes that 
converting casual riders into annual members can significantly enhance their 
profitability. To devise effective marketing strategies for this conversion, the 
case study will delve into questions that probe how these two user groups use 
Cyclistic bikes differently, why casual riders might opt for annual memberships, 
and how digital media can facilitate this conversion.

This study is divided into six stages: Ask, Prepare, Process, Analyze, Share, 
and Act. Each section will come with additional detail and build on the previous 
section. We hope you enjoy reading this analysis as much as we enjoyed making
it. All work was done with the R Language [@rbase] in R Studio [@rstudio].


# Ask
The primary business question we aim to address is: How do annual members and 
casual riders use Cyclistic bikes differently? This question is fundamental to 
developing a marketing strategy that can effectively convert casual riders into 
annual members.

To answer this question, we will delve into the following topics:

1. *Duration and frequency of rides:* We will analyze data to understand the 
typical length and frequency of rides for both user groups. Are there noticeable 
differences between casual riders and annual members? Are members using the 
bikes more frequently or for longer duration?

2. *Purpose of rides:* We aim to determine the typical purpose of rides for both 
groups. Do casual riders primarily use the bikes for leisure, while members use 
them more for daily commuting? Or is there an overlap?

3. *Ride timings:* We'll investigate if there's a difference in the time of day, 
day of the week, or month of the year when the two user groups typically use the 
bikes.

4. *Bike preferences:* Given that Cyclistic offers various bike options, we will 
examine if there's a pattern in the type of bike preferred by the two user 
groups.

Answering these questions will provide insights into the riding patterns of the 
two groups, and will be instrumental in forming the foundation for a marketing 
strategy aimed at converting casual riders into annual members. The strategy 
will be developed in alignment with these insights, addressing the unique needs 
and behaviors of casual riders to motivate their transition to becoming annual 
members.

Cyclistic's collected data forms the backbone of this analysis. For 
transparency, we have included the source code in the paper's Appendix. We 
strive to communicate our findings clearly, pairing concise language with 
comprehensible visualizations and straightforward recommendations.

# Prepare  

The data for this analysis is publicly available and located on Divvy's Amazon 
S3 storage. The data is titled as "Divvy Data." [@DivvyData2023]

The data is organized in an anonymized format, detailing each trip's start and 
end day/time, start and end stations, and rider type (Member or Casual). Trips 
taken by staff and any trips under 60 seconds in length have been excluded from 
this dataset.

Given the anonymization and rigorous data handling practices outlined by Divvy, 
we can consider the data to be Reliable, Original, Comprehensive, Current, and 
Cited (ROCCC). However, some level of bias could be inherent as the data only 
includes riders who choose to use Divvy's services, and not all potential or 
actual cyclists in Chicago. Also, the credibility of the dataset is tied to 
Divvy's data collection and handling processes.

The data is provided according to the Divvy Data License Agreement. 
Accessibility is facilitated through public download links, ensuring the data 
can be freely used for analysis while respecting privacy through anonymization. 
Security of the data in our analysis will be ensured by following best 
practices, such as secure data storage and handling.

Data integrity has been verified through initial exploratory data analysis to 
check for missing values, inconsistencies, or obvious outliers. Any issues found 
during this process will be documented and addressed appropriately.

This dataset is instrumental in answering our primary business question — 
understanding how annual members and casual riders use Cyclistic bikes 
differently. Each trip's information, coupled with the type of rider, will offer 
insights into the usage patterns of the two user groups.

Potential problems with the data could include missing values, inconsistent data 
entries, or anomalies. These will be explored and addressed during the data 
cleaning and preprocessing stages of our analysis. If severe issues are 
encountered, they will be documented and taken into consideration when 
interpreting the analysis results.

## Structure of the Data
We used the str() function in R to give us a detailed structure of the data.  

```{r, echo = TRUE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}

# examine the structure of the data
str(combined_data_df)

```

Here are some interesting observations from the output:  

* __Data size:__ The dataframe contains over 6.23 million records, spread across 13 
variables.  

* __Data types:__ The variables in the dataframe are of various types, including 
character (e.g., ride_id, start_station_name, start_station_id, etc.), 
factor (e.g., rideable_type, member_casual), numeric (e.g., start_lat, 
start_lng, end_lat, end_lng), and POSIXct (e.g., started_at, ended_at). This 
variety indicates that the dataset is rich and contains different kinds of 
information.  

* __Rideable types:__ There are three levels of rideable_type: classic_bike, 
docked_bike, and electric_bike.  

* __Member categories:__ There are two levels of member_casual: casual and member.  

* __Geographic coordinates:__ The data contains latitude and longitude information 
for both the start and end locations of rides. This information could be used to 
plot geographic routes or to analyze usage patterns across different locations.  

* __Time information:__ The dataset includes when each ride started and ended. This 
information can be used to analyze the temporal patterns of bike usage.

* __Station information:__ The data includes detailed information about the start 
and end stations, including their names and IDs.

```{r, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, appendix = TRUE}



```

\newpage
# Appendix

## References  

<div id="refs"></div>  

## Source Code  

```{r ref.label = knitr::all_labels(appendix == TRUE), echo=TRUE, eval=FALSE}
